Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

### 2018-12-07 paper/sections/methodology.tex
All corrections completed

Probe requests are a special type of management packet broadcast by Wi-Fi
enabled devices as part of their various functions such as scanning for
available APs and quick geolocation by triangulation based known APs, etc,.
These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer,
type or model of the devices, although there is some variation in the frequency
and the content of the information transmitted through them. In some cases, such
as Android devices, these are broadcast even when the Wi-Fi functionality has
been turned off by the user so that the device can immediately connect to
networks when the functionality is switched back on. Since some devices even use
the probe requests as a less accurate form of localisation, they continuously
send probe requests when Wi-Fi has been switched off. Thus, these signals can be
used to reliably identify the presence of Wi-Fi enabled mobile devices. Being a
first step of connection initiated by the mobile device, these packets have
information regarding the characteristics of the mobile device itself. Some of
the key information we can infer from these requests are,

@@

After collecting data, we began estimating the footfall or pedestrian
activity from them by identifying the following potential uncertainties arising
from our data collection method.

\begin{enumerate} 
\item 
\textbf{Background noise} - since the extent to which Wi-Fi signals travel
		differs subject to various factors such as interference and humidity, it
		is close to impossible to restrict our data collection to a finite area
		of interest. This can lead to a significant background noise at certain
		locations. E.g. a phone shop or a bus stop located next to the study
		area can increase the number of probe requests received by the sensor.
		It is important to note that this method may not work effectively on
		study locations with complex configurations such as the source of noise
		and the area of study being located at the same distance from the
		sensor. This aspect is explored in detail in the broader case study in
		the following sections.
\item 
\textbf{MAC randomisation} - The mobile devices in recent years have been using
		randomised 'local' MAC addresses for probe requests to protect the users
		from being tracked. This makes it impossible to tell if the probe
		requests are being sent by the same mobile device which is being
		stationed next to the sensor. This along with the previous problem can
		further increase the magnitude of error by several fold.
\item
\textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely
		across geography and demography, we cannot assume that every mobile
		device translates to one pedestrian footfall. In addition to this, there
		is a long term overall increase in mobile ownership which may lead to
		the number of probe requests collected overtime.
\end{enumerate}

@@

One of the clues that we can use to estimate the distance between the mobile
device and the sensor is the strength of the signal received by the sensor. The
obvious approach here is to try and establish a relationship between the signal
strength and distance first and use this to filter out the unwanted probe
requests. This approach was found not to be feasible since the decay of signal
strength with distance is not always constant. It varies with atmospheric
conditions, presence of obstructions between the source and target, the nature
of these obstructions and the strength (power level) of the source. This
severely limits our ability in establishing a simple conversion between reported
signal strength and distance. There is a need for a method which takes in to
account these variables across various locations.

We assume that in configurations where a specific source of background
noise is at a constant distance, there must be a distinct pattern in the number
of probe requests reporting signal strength corresponding to that distance. For
example, if there is a phone shop next to our sensor where hundreds of phones
regularly send probe requests there should be a sharp rise of number of probe
requests with reported signal strength corresponding to the distance between the
sensor and the phone shop irrespective of the local conditions as shown in
Figure \ref{methodology_schematic}. We could identify these breaks in the data
using traditional one dimensional clustering algorithms such as `jenks natural
breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
only looking for the break in the data and not for absolute values, the
methodology should apply for all the variations due micro site conditions thus
reducing the overall noise in the collected data.

@@

Since our primary unique identifier - MAC address, is being anonymised by new
devices, we need to find other information present in the probe request for use
as a unique identifier. Obvious approach here is to establish a factor of
randomisation and adjust the counts for these probe requests based on this
factor. We found this approach to be not feasible since the proportion of
devices which randomise the MAC addresses increases over time. There is also a
wide variation in the frequency at which the devices randomise the MAC addresses
and the method used for the process. This lead us to look for a more
generalisable approach which is independent of the device model.

From our initial look at the data we found that OUI and the sequence
number of the packet is the most promising information to achieve this.
First we divide our dataset into sets of probe requests with randomised and
non-randomised MAC addresses by looking at the second character of the vendor
part of the MAC address and if it is E, A, 2 or 6, then these addresses are
identified to be randomised. We keep the MAC address as the unique
identifier for the non-randomised requests and further divide the randomised ones in to
sub categories based on their OUI. We then  
identify unique mobile devices from within these sets and assign a unique identifier to each device.

The proposed algorithm creates a graph where the probe requests represented the
nodes, and links are created between them based on the following rules:

@@

The nodes were then assigned a unique id based on the unique connected component
they belong to as shown in Figure \ref{methodology_schematic}. This unique
identifier is used in the place of MAC address for aggregation for the
anonymised probe requests. Though the recycling of sequence number after 4096
leads to multiple unique ids being reported from a single device from a sample
experiment done on randomised probe requests sent by "Google" devices we found
that the sequence number is reset for around 0.05\% of them. We hence assume
this to be inconsequential.

@@

Since mobile device ownership is an external uncertainty to our study and could
arise from variety of spatio - temporal and demographic factors, we propose to
solve this using a sample count done manually at each location.  We can then
calculate an  adjustment factor or an offset for each location by comparing the
sensor based counts and ground truth and in turn it can be used to adjust the
data reliably to reflect the ground truth in absolute numbers for the future.
This calibration can be carried out periodically at these locations to improve
the quality of the estimation.

>>>>

Probe requests are a special type of management packet broadcast by Wi-Fi
enabled devices as part of their various functions such as scanning for
available APs and quick geolocation by triangulation based known APs, etc.
These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer,
type or model of the devices, although there is some variation in the frequency
and the content of the information transmitted through them. In some cases, such
as Android devices, these are broadcast even when the Wi-Fi functionality has
been turned off by the user so that the device can immediately connect to
networks when the functionality is switched back on. Since some devices even use
the probe requests as a less accurate form of localisation, they continuously
send probe requests when Wi-Fi has been switched off. Thus, these signals can be
used to reliably identify the presence of Wi-Fi enabled mobile devices. Being a
first step of connection initiated by the mobile device, these packets have
information regarding the characteristics of the mobile device itself. Some of
the key information we can infer from these requests are,

@@

After collecting data, we began estimating the footfall or pedestrian activity
from them by identifying the following potential uncertainties arising from our
data collection method:

\begin{enumerate} 
\item 
\textbf{Background noise} - since the extent to which Wi-Fi signals travel
        differs subject to various factors such as interference and humidity, it
        is close to impossible to restrict our data collection to a finite area
        of interest. This can lead to a significant background noise at certain
        locations. For example, a phone shop or a bus stop located next to the
        study area can artificially increase the number of probe requests
        received by the sensor.  It is important to note that this method may
        not work effectively on study locations with complex configurations such
        as the source of noise and the area of study being located at the same
        distance from the sensor. This aspect is explored in detail in the
        broader case study in the following sections.
\item 
\textbf{MAC randomisation} - mobile devices in recent years have been using
        randomised `local' MAC addresses for probe requests to protect the users
        from being tracked. This makes it impossible to tell if the probe
        requests are being sent by the same mobile device. This along with the
        previous problem can further increase the magnitude of error by several
        fold.
\item
\textbf{Mobile ownership} - since the rate of mobile ownership can vary widely
		across geography and demography, we cannot assume that every mobile
		device translates to one pedestrian footfall. In addition to this, there
		is a long term overall increase in mobile ownership which may affect 
		the number of probe requests collected overtime.
\end{enumerate}

@@

One of the clues that we can use to estimate the distance between the mobile
device and the sensor is the strength of the signal received by the sensor. The
obvious approach was to first try and establish a relationship between the
signal strength and distance and to use this to filter out the unwanted probe
requests. However this approach was found to not be feasible, since the decay of
signal strength with respect to distance is not always constant. For instance,
signal strength varies with atmospheric conditions, the presence of obstructions
between the source and the target, the nature of these obstructions, and the
strength (power level) of the source. This severely limits our ability to
establish a simple conversion between reported signal strength and distance. As
such, there was a need for a method which takes in to account all of these
variables across the various locations.

We assumed that, in configurations where a specific source of background noise
was at a constant distance, there must be a distinct pattern in the number of
probe requests reporting signal strength corresponding to that distance. For
example, if there was a phone shop next to our sensor where hundreds of phones
regularly sent probe requests, there should be a sharp rise in the of number of
probe requests with reported signal strength corresponding to the distance
between the sensor and the phone shop, irrespective of the local conditions as
shown in Figure \ref{methodology_schematic}. We could identify these breaks in
the data using traditional one dimensional clustering algorithms such as `jenks
natural breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since
we were only looking for the break in the data and not for absolute values, the
methodology should apply for all the variations due to micro site conditions
reducing the overall noise in the collected data.

@@

Since our primary unique identifier - MAC addresses are being anonymised by new
devices, we needed to find other information present in the probe requests for
use as a unique identifier. The obvious approach was to establish a factor of
randomisation, and adjust the counts for the probe requests based on this
factor. We found this approach to not be feasible since the proportion of
devices which randomise the MAC addresses increased over time. There was also a
wide variation in the frequency at which the devices randomised the MAC
addresses and the method used for the process. This led us to look for a more
generalisable approach which was independent of the device model.

From our initial look at the data, we found that OUI and the sequence number of
the packet was the most promising information to achieve this.  First we divided
our dataset into sets of probe requests with randomised and non-randomised MAC
addresses by looking at the second character of the vendor part of the MAC
address; if it was E, A, 2 or 6, then those addresses were identified to be
randomised. We kept the MAC address as the unique identifier for the
non-randomised requests and further divided the randomised ones in to
sub-categories based on their OUI. We then  identified unique mobile devices
from within those sets, and assigned a unique identifier to each device.

The proposed algorithm created a graph where the probe requests represented the
nodes; links were created between them based on the following rules:

@@

The nodes were then assigned a unique ID based on the unique connected component
they belonged to as shown in Figure \ref{methodology_schematic}. This unique
identifier was used in the place of MAC addresses for aggregation of the
anonymised probe requests. Although the recycling of sequence numbers after 4096
led to multiple unique IDs being reported from a single device, a sample
consisting of all randomised probe requests sent by "Google" devices showed that
only 0.05\% of the sample had their sequence number reset. This led assume this
to be inconsequential.

@@

Since proportions of mobile device ownership was an external uncertainty to our
study and could arise from variety of spatio - temporal and demographic factors,
we aimed to solve this by using a manual sample count at each location. We then
calculated an adjustment factor, or an `offset' for each location by comparing
the sensor-based counts and ground truth. In turn it was then used to adjust the
data reliably to reflect the ground truth in absolute numbers.  This calibration
can be carried out periodically at these locations to improve the quality of the
estimation.
